Elephant Seal Age JAE Intraspecific Variation

### 2023-12-08 paper/paperv3.qmd
Fix quadratic fits, add year lines to fig 1b

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987)

@@

breeding_mod <- glmer(observed_int ~ poly(age, 2) + (1 | animalID) + (1 | year),
                      data = sealdat, 
                      family = "binomial", 
                      control = glmerControl(optimizer = "bobyqa"))

breed_pred <- ggpredict(breeding_mod, terms = c("age [all]"))

@@

@@

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(x, ymin = conf.low, ymax = conf.high),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(x, predicted),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) + 
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

@@

##NOTE: Need to take out the instances where recruitment > survival

surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  filter(pupsex == "F",
         pupyearborn < 2016) %>% 
  mutate(pup_survived_dbl = ifelse(pup_survived, 1, 0)) %>% 
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>% 
  filter(n_pups >= 5)
pup_surv_model <- glmer(pup_survived_dbl ~ poly(age, 2) + (1 | animalID) + (1 | year),
                        data = surv_recruit_data, 
                        family = "binomial")
pup_surv_pred <- ggpredict(pup_surv_model, terms = c("age [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x, 
            metric = "surv", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
pup_recruit_model <- glmer(puprecruited ~ poly(age, 2) + (1 | animalID) + (1 | year),
                           data = surv_recruit_data, 
                           family = "binomial")
pup_recruit_pred <- ggpredict(pup_recruit_model, terms = c("age [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x, 
            metric = "recruit", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
surv_recruit_pred <- rbind(pup_surv_pred, pup_recruit_pred) %>% 
  mutate(metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

@@

#linear
pup_sex_mod <- glmer(is_male ~ age + (1 | animalID) + (1 | year),
                     data = pup_sex_data, 
                     family = "binomial")

@@

seal_phenology <- sealdat %>% 
  filter(observed == "Breeder") %>% 
  mutate(BA = firstobsbreeddoy, 
         BD = firstobsbreeddoy + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

@@

#NOTE: Both BA and BD have some points that are really late (above 300 doy) that need fixing
BA_age <- ggplot(seal_phenology, aes(age, BA)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 150)) + 
  theme_classic() + 
  labs(y = "Breeding arrival (DOY)", x = "Maternal age (Years)")

BD_age <- ggplot(seal_phenology, aes(age, BD)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 150)) + 
  theme_classic() + 
  labs(y = "Breeding departure (DOY)", x = "Maternal age (Years)")

MA_age <- ggplot(seal_phenology, aes(age, MA)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "#829BD1") + 
  scale_y_continuous(limits = c(50, 200)) + 
  theme_classic() + 
  labs(y = "Molt arrival (DOY)", x = "Maternal age (Years)")

MD_age <- ggplot(seal_phenology, aes(age, MD)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "#829BD1") + 
  scale_y_continuous(limits = c(50, 200)) + 
  theme_classic() + 
  labs(y = "Molt departure (DOY)", x = "Maternal age (Years)")

plot_grid(BA_age, BD_age, MA_age, MD_age, nrow = 2, ncol = 2)
```

>>>>

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(across(c(animalID, year), factor),
         age10 = age / 10)

@@

breeding_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = sealdat, 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

breed_pred <- ggpredict(breeding_mod, terms = "age10 [all]") %>% 
  as_tibble() %>% 
  mutate(age = x * 10)

ranef_year <- ranef(breeding_mod)$year %>% 
  as_tibble(rownames = "year") %>% 
  arrange(`(Intercept)`) %>% 
  slice(1, floor(nrow(.) * c(0.25, 0.5, 0.75, 1))) %>% 
  expand_grid(age10 = unique(sealdat$age10)) %>% 
  mutate(predicted = predict(breeding_mod, newdata = ., re.form = ~ (1 | year), type = "response"),
         age = age10 * 10)

@@

breed_mod_summary

@@

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = year),
            ranef_year,
            inherit.aes = FALSE,
            alpha = 0.2) +
  geom_text(aes(x = 22.1, predicted, label = year),
            filter(ranef_year, age == 22),
            inherit.aes = FALSE,
            hjust = 0,
            alpha = 0.5,
            size = 3) +
  geom_line(aes(age, predicted),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) + 
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

@@

##NOTE: Need to take out the instances where recruitment > survival
#what does ggpredict do with random effects? 
#thin gray lines for the different year effects? show the interannual variation (5 that include extreme ones, just something to illustrate the variation of year)

surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  filter(pupsex == "F",
         pupyearborn < 2016) %>% 
  mutate(pup_survived_dbl = ifelse(pup_survived, 1, 0)) %>% 
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>% 
  filter(n_pups >= 5)
pup_surv_model <- glmer(
  pup_survived_dbl ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = surv_recruit_data, 
  family = "binomial"
)
summary(pup_surv_model)
pup_surv_pred <- ggpredict(pup_surv_model, terms = c("age10 [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x * 10, 
            metric = "surv", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
pup_recruit_model <- glmer(
  puprecruited ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = surv_recruit_data, 
  family = "binomial"
)
summary(pup_recruit_model)
pup_recruit_pred <- ggpredict(pup_recruit_model, terms = c("age10 [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x * 10, 
            metric = "recruit", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
surv_recruit_pred <- rbind(pup_surv_pred, pup_recruit_pred) %>% 
  mutate(metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

@@

#linear
pup_sex_mod <- glmer(is_male ~ age + (1 | animalID) + (1 | year),
                     data = pup_sex_data, 
                     family = "binomial")
summary(pup_sex_mod)

@@

seal_phenology <- sealdat %>% 
  filter(observed == "Breeder") %>% 
  mutate(BA = ifelse(firstobsbreeddoy > 200, firstobsbreeddoy - 365, firstobsbreeddoy), 
         BD = BA + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

@@

#NOTE: Both BA and BD have some points that are really late (above 300 doy) that need fixing. Maybe make boxplots also..?
seal_phenology %>% 
  rename(`Breeding\narrival` = BA,
         `Breeding\ndeparture` = BD,
         `Molt\narrival` = MA,
         `Molt\ndeparture` = MD) %>% 
  pivot_longer(`Breeding\narrival`:`Molt\ndeparture`, 
               names_to = "event", 
               values_to = "DOY") %>% 
ggplot(aes(y = DOY)) +
  geom_boxplot(aes(factor(age))) +
  geom_smooth(aes(as.integer(factor(age))), method = "lm", formula = y ~ x, se = TRUE) +
  facet_wrap(~ event, nrow = 2, scales = "free_y") +
  labs(x = "Age (years)", y = "Day of year") +
  theme_classic()
```
